The \emph{FThresholds} package for \emph{Macaulay2}

### 2019-06-04 Fsing/FThresholdsPaper/FThresholds.tex

\documentclass{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,setspace}
%\usepackage{mathtools}
\RequirePackage[dvipsnames,usenames]{xcolor}

@@

\usepackage{mabliautoref}
\usepackage{colonequals}
\frenchspacing
\input{kmacros3.sty}
\usepackage{stmaryrd}

\usepackage{verbatim}
\usepackage{enumerate}

\usepackage[normalem]{ulem}
%\usepackage{marginnote}

\DeclareMathOperator{\HH}{H}
\newcommand{\fram}{\mathfrak{m}}
%\DeclareMathOperator{\Image}{image}

\DefineVerbatimEnvironment%
  {MyVerbatim}{Verbatim}
  {formatcom=\color{Violet}}

\renewcommand{\geq}{\geqslant}
\renewcommand{\leq}{\leqslant}
\renewcommand{\ge}{\geqslant}
\renewcommand{\le}{\leqslant}

\newtheorem*{caveat}{Caveat}

\begin{document}
\title{The  \emph{TestIdeals} package for \emph{Macaulay2}}
\author[Alberto F.\ Boix et al.]{Alberto F.\ Boix}
\address{Department of Mathematics, Ben-Gurion University of the Negev, Beer-Sheva 8410501, Israel}
\email{fernanal@post.bgu.ac.il}
\thanks{A.F.\,Boix was supported by Israel Science Foundation (grant No. 844/14) and Spanish Ministerio de Econom\'ia y Competitividad MTM2016-7881-P}

@@

\author[]{Zhibek Kadyrsizova}
\address{School of Science and Technology, Nazarbayev University, Astana, 010000, Republic of Kazakhstan}
\email{zhibek.kadyrsizova@nu.edu.kz}
\thanks{Z. Kadyrsizova was partially supported by NSF DMS \#1401384 and the Barbour Scholarship at the University of Michigan.}

\author[]{Mordechai Katzman}
\address{Department of Pure Mathematics, University of Sheffield, Sheffield S37RH, United Kingdom}
\email{M.Katzman@sheffield.ac.uk}

\author[]{Sara Malec}
\address{Department of Mathematics, Hood College, Frederick, MD 21701}
\email{malec@hood.edu}

\author[]{Marcus Robinson}
\address{Department of Mathematics, University of Utah, Salt Lake City, UT~84112, USA}
\email{robinson@math.utah.edu}

@@

\author[]{Daniel Smolkin}
\address{Department of Mathematics, University of Utah, Salt Lake City, UT~84112, USA}
\email{smolkin@math.utah.edu}
\thanks{D.~Smolkin was supported by NSF RTG Grant DMS \#1246989, NSF CAREER Grant DMS \#1252860/1501102, and NSF FRG Grant DMS \#1265261/1501115.}

@@

\begin{abstract}
	This note describes a \emph{Macaulay2} package for computations of $F$(-pure) thresholds in characteristic $p > 0$ commutative algebra.  
\end{abstract}

@@

There is one more option that deserves mentioning, {\tt ContainmentTest}.  There are two ways to check whether $f^a$ (or $I^a$) is contained in $J^{[p^e]}$.  In general it is true that
\[
I^a \subseteq J^{[p^e]} \text{ $\Leftrightarrow$ } (I^a)^{[1/p^e]} \subseteq J
\]
where $\bullet^{[1/p^e]}$ is defined as in the {\tt TestIdeals} package.  These are controlled by {\tt ContainmentTest => StandardPower} and {\tt ContainmentTest => FrobeniusRoot} respectively.
For principal $I$, the command on the right is generally faster (and so {\tt FrobeniusRoot} is the default strategy for computing $\nu$).  For non-principal $I$, frequently the command on the left is quicker, and so {\tt StandardPower} is default strategy.

@@

\section{\tt fpt}
\label{sec.FPT}

The core function in this package is the {\tt fpt} function.  Given an element $f \in R = k[x_1, \dots, x_n]$, the {\tt fpt} function tries to compute the $F$-pure threshold of $f$.  It does this in the following way:

\section{}

\bibliographystyle{skalpha}
\bibliography{MainBib}

>>>>

\documentclass{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,setspace}
\RequirePackage[dvipsnames,usenames]{xcolor}

@@

\usepackage{mabliautoref}
\usepackage{colonequals}
\frenchspacing
\input{kmacros3.sty}

\usepackage{verbatim}
\usepackage{enumerate}
\usepackage[normalem]{ulem}
%\usepackage{marginnote}
\newcommand{\fram}{\mathfrak{m}}
\DefineVerbatimEnvironment%
{MyVerbatim}{Verbatim}
{formatcom=\color{Violet}}

\begin{document}
\title{The  \emph{FThresholds} package for \emph{Macaulay2}}
%\author[Alberto F.\ Boix et al.]{Alberto F.\ Boix}
%\address{Department of Mathematics, Ben-Gurion University of the Negev, Beer-Sheva 8410501, Israel}
%\email{fernanal@post.bgu.ac.il}
%\thanks{A.F.\,Boix was supported by Israel Science Foundation (grant No. 844/14) and Spanish Ministerio de Econom\'ia y Competitividad MTM2016-7881-P}

@@

%\author[]{Zhibek Kadyrsizova}
%\address{School of Science and Technology, Nazarbayev University, Astana, 010000, Republic of Kazakhstan}
%\email{zhibek.kadyrsizova@nu.edu.kz}
%\thanks{Z. Kadyrsizova was partially supported by NSF DMS \#1401384 and the Barbour Scholarship at the University of Michigan.}

%\author[]{Mordechai Katzman}
%\address{Department of Pure Mathematics, University of Sheffield, Sheffield S37RH, United Kingdom}
%\email{M.Katzman@sheffield.ac.uk}

%\author[]{Sara Malec}
%\address{Department of Mathematics, Hood College, Frederick, MD 21701}
%\email{malec@hood.edu}

%\author[]{Marcus Robinson}
%\address{Department of Mathematics, University of Utah, Salt Lake City, UT~84112, USA}
%\email{robinson@math.utah.edu}

@@

%\author[]{Daniel Smolkin}
%\address{Department of Mathematics, University of Utah, Salt Lake City, UT~84112, USA}
%\email{smolkin@math.utah.edu}
%\thanks{D.~Smolkin was supported by NSF RTG Grant DMS \#1246989, NSF CAREER Grant DMS \#1252860/1501102, and NSF FRG Grant DMS \#1265261/1501115.}

@@

\begin{abstract}
	This note describes a \emph{Macaulay2} package for computations of $F$-pure thresholds, more general $F$-thresholds, and other related numerical invariants, in prime characteristic commutative algebra.  
\end{abstract}

@@

There is one more option that deserves mentioning, {\tt ContainmentTest}.  There are two ways to check whether $f^a$ (or $I^a$) is contained in $J^{[p^e]}$.  In general it is true that
\[
I^a \subseteq J^{[p^e]} \text{ $\Leftrightarrow$ } (I^a)^{[1/p^e]} \subseteq J
\]
where $\bullet^{[1/p^e]}$ is defined as in the {\tt TestIdeals} package.  Which of these two strategies to use is controlled by the option {\tt ContainmentTest => StandardPower} and {\tt ContainmentTest => FrobeniusRoot} respectively.
For principal $I$, the command on the right is generally faster (and so {\tt FrobeniusRoot} is the default strategy for computing $\nu$).  For non-principal $I$, frequently the command on the left is quicker, and so {\tt StandardPower} is default strategy.

@@

\newpage
\section{The {\tt fpt} function}
\label{sec.FPT}

The core function in this package is the {\tt fpt} function.  Given a polynomial $f$ with coefficients in a finite field, the function {\tt fpt} attempts to find the exact value for the $F$-pure threshold of $f$ at the origin, and returns that value, if possible.  Otherwise, it returns lower and upper bounds for the $F$-pure threshold, as demonstrated below.

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i1 : ZZ/5[x,y,z];

i2 : fpt( x^3 + y^3 + z^3 + x*y*z )

     4
o2 = -
     5

o2 : QQ

i3 : fpt( x^5 + y^6 + z^7 + (x*y*z)^3 )

      1  2
o3 = {-, -}
      3  5

o3 : List
\end{MyVerbatim}
}

             If the option {\tt UseSpecialAlgorithms} is set to {\tt true} (the default value), then {\tt fpt} first checks whether $f$ is diagonal polynomial, a binomial, or a form in two variables, respectively.
             If it is one of these, algorithms of Hern\'andez, or Hern\'andez and Teixeira, {\color{red}  References?} are executed to compute the $F$-pure threshold of $f$.

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i4 : fpt( x^17 + y^20 + z^24 ) -- a diagonal polynomial

      94
o4 = ---
     625

i5 : fpt( x^2*y^6*z^10 + x^10*y^5*z^3 ) -- a binomial

      997
o5 = ----
     6250

i6 : ZZ/5[x,y];

i7 : fpt( x^2*y^6*(x + y)^9*(x + 3*y)^10 ) -- a binary form in two variables

      5787
o7 = -----
     78125

o7 : QQ
\end{MyVerbatim}
}

            The computation of the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and can sometimes hang when attempting that factorization. For this reason, when a factorization is already known, the user can pass to {\tt fpt} a list containing all the pairwise prime linear factors of $f$ and a list containing their respective multiplicities.

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i8 : L = {x, y, x + y, x + 3*y};

i9 : m = {2, 6, 9, 10};

i10 : fpt(L, m)==o7

o10 = true
\end{MyVerbatim}
}

The behavior of the {\tt fpt} function is more involved when no special algorithm is available, or {\tt UseSpecialAlgorithms} is set to {\tt false}.  In the remainder of this section, we describe the behavior of the {\tt fpt} function in this case, as well as the roles of the options {\tt DepthOfSearch}, and {\tt Attempts}, both nonnegative integers.

            When no special algorithm is available or , then function {\tt fpt} computes {\tt nu}$(e,f)$, where $e$ is the value of the option {\tt DepthOfSearch}, which conservatively defaults to $1$.
            At this point, we know that the $F$-pure threshold of $f$ lies in the closed interval {\color{red} Elaborate here} [$\nu/(p^e-1),(\nu+1)/p^e$], and the subroutine {\tt guessFPT} is called to make some "educated guesses" in an attempt to find the $F$-pure threshold, or at least narrow down the above interval.
	    The number of "guesses" is controlled by the option {\tt Attempts}, which conservatively defaults to 3.
            If {\tt Attempts} is set to 0, {\tt guessFPT} is bypassed.
            If  {\tt Attempts} is set to at least 1, then a first check is run to verify whether the right-hand endpoint $(\nu+1)/p^e$ of the above interval is the $F$-pure threshold.

\begin{example}
\[             f = x^2*(x + y)^3*(x + 3*y^2)^5;
            fpt( f, Attempts => 0 ) -- a bad estimate
            fpt( f, Attempts => 0, DepthOfSearch => 3 ) -- a better estimate
            fpt( f, Attempts => 1, DepthOfSearch => 3 ) -- the right-hand endpoint (nu+1)/p^e is the fpt \] 
\end{example}

            If  {\tt Attempts} is set to at least 2 and the right-hand endpoint $(\nu+1)/p^e$ is not the $F$-pure threshold, a second check is run to verify whether the left-hand endpoint $\nu/(p^e-1)$ is the $F$-pure threshold.

\begin{example}
\[
	    f = x^6*y^4 + x^4*y^9 + (x^2 + y^3)^3;
            fpt( f, Attempts => 1, DepthOfSearch => 3 )
            fpt( f, Attempts => 2, DepthOfSearch => 3 ) -- the left-hand endpoint nu/(p^e-1) is the fpt
\]
\end{example}

            If neither endpoint is the $F$-pure threshold, and {\tt Attempts} is set to more than 2, additional checks are performed at numbers in the interval.
            A number in the interval with minimal denominator is selected, and {\tt compareFPT} is used to test that number.
            If that ``guess" is correct, its value is returned; otherwise, the information returned by {\tt compareFPT} is used to narrow down the interval, and this process is repeated as many times as specified by {\tt Attempts}.

\begin{example}
\[             f = x^3*y^11*(x + y)^8*(x^2 + y^3)^8;
            fpt( f, DepthOfSearch => 3, Attempts => 2 )
            fpt( f, DepthOfSearch => 3, Attempts => 3 ) -- an additional check sharpens the estimate
            fpt( f, DepthOfSearch => 3, Attempts => 4 ) -- and one more finds the exact answer
\]
\end{example}

            If {\tt guessFPT} is unsuccessful and {\tt UseFSignature} is set to {\tt true}, the fpt function proceeds to use the convexity of the $F$-signature function and a secant line argument to attempt to narrow down the interval bounding the $F$-pure threshold.

\begin{example}
\[             f = x^5*y^6*(x + y)^9*(x^2 + y^3)^4;
            fpt( f, DepthOfSearch => 3 )
            fpt( f, DepthOfSearch => 3, UseFSignature => true )
            numeric ooo
            numeric ooo -- UseFSignature sharpened the estimate a bit \]
\end{example}

            When {\tt FRegularityCheck} is set to {\tt true} and no exact answer has been found, a final check is run (if necessary) to verify whether the final lower bound for the $F$-pure threshold is the exact answer.

\begin{example}
\[             f = (x + y)^4*(x^2 + y^3)^6;
            fpt( f, Attempts => 2, DepthOfSearch => 3 )
            fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true ) -- using FSignatures the answer improves a bit
	    fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true, FRegularityCheck => true ) -- FRegularityCheck finds the answer \] 
\end{example}

            The computations performed when {\tt UseFSignature} and {\tt FRegularityCheck} are set to {\tt true} are often slow, and often fail to improve the estimate, and for this reason, these options should be used sparingly.
            It is often more effective to increase the values of {\tt Attempts} or {\tt DepthOfSearch}, instead.

\begin{example}            
\[            f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;
	    timing numeric fpt( f, DepthOfSearch => 3, UseFSignature => true, FRegularityCheck => true )
            timing numeric fpt( f, Attempts => 5, DepthOfSearch => 3 ) -- a better answer in less time
            timing fpt( f, DepthOfSearch => 4 ) -- the exact answer in even less time
\]
\end{example}

            As seen in several examples above, when the exact answer is not found, a list containing the endpoints of an interval containing the $F$-pure threshold of $f$ is returned.
            Whether that interval is open, closed, or a mixed interval depends on the options passed; if the option {\tt Verbose} is set to {\tt true}, the precise interval will be printed.

\begin{example}
 \[           f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;
            fpt( f, DepthOfSearch => 3, UseFSignature => true, Verbose => true )
 \]
  \end{example}

\newpage
\bibliographystyle{skalpha}
\bibliography{MainBib}
